Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-14 aistats.tex
Clean and add links to appendix

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$. %Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[e.g. \S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
Hence the fundamental problem:
%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We know asymptotic answers.
We also know simple upper bounds if $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
or close to quadratic (see \S\ref{ssec:quadratic}).
But in general, this problem is difficult. In this paper, we show some partial solutions, and give some ideas on how to solve it.
%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
%RLP poroblem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.

@@

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
A convergence rate for this algorithm would translate into the upper bound we seek.
%From an optimization perspective, this problem is interesting, as it is often non-smooth (e.g. not well approximated by quadratics), but benefits from \emph{relative} smoothness properties \citep{birnbaum2011distributed}.
Conversely we can see this problem as a very special case to study SMD, revealing the limits of current analysis.

@@

%Or it could help developing accelerated (stochastic?) algorithms for barrier objectives  overcoming the limitations of non-Euclidean geometry \citep{dragomir2021optimal}. \rlp{Fred I need your feedback on that one.}

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance})
	\item We use a connection between~\eqref{problem} and optimization to highlight areas for progress in the stochastic optimization of barrier objectives (\S\ref{sec:optimization}).
\end{itemize}

@@

\paragraph{Optimization}
We can rewrite~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
}
which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
\emph{MAP is an instantiation of SMD.}
Note that this analogy does not hold for the MLE %($n_0=0$)
as it does not have any initialization to start from.
\fdk{It does, you just forget the initial condition after the first step with step-size 1?}
\rlp{to discuss.}

@@

@@

In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
\begin{theorem}[MLE Tight Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound is asymptotically tight.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
We get a similar bound for the multivariate generalization :
the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (see appendix).

@@

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
\rlp{upper bound and compare the symmetrized bregman instead?}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$. This correlates well with numerical observations (see appendix).
Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.

@@

\subsection{Asymptote}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now review classical asymptotic results.
We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\alignn{
    \bregmanconj(\m^* ; \m)
    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify
\begin{align}
	\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
See proof in the Appendix.
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
FromTaylor expansion~~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, like self-concordance of $\conj$ \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
Proof in the appendix
\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$ (see appendix).
This means that on average, we need much more samples than the dimension to reach the quadratic regime.

@@

\newpage
\bibliographystyle{apalike}
\bibliography{references} % Removed the .bib as this mess up with the compiler

\clearpage
\section{Group Meeting}
\begin{enumerate}
	\item new plots : mirror map, and gaussian trajectories. Add trajectories to paper, with background level set, and new legend : primal/dual expectation
	\item the MLE is biased according to our definition. See plots. Does this definition of bias make sense ? An unbiased estimator's primal mean would be equal to the optimum, vs the dual mean.  Remark : the MLE is biased for standard parameters.
	\item exciting pap er\citet{kakade2010learning}
	\begin{enumerate}
		\item assumption in the spirit of infinite order self-concordance only at the optimum
		\item After a burn-in phase, the loss behaves quadratically. No precise description of the burn-in length.
		\item \TODO{it is worth writing down until the end what we are obtaining with self-concordance.}
		\item idea 1 : try to express some of their results with the mean parameter instead.
		\item idea 2 : using their assumption on the mirror map, perform a Newton kind of analysis for SMD. (exploiting the link MAP -> optimization). How do we prove the linear convergence of Newton to the quadratic regime ? That is the problem at stake here.
	\end{enumerate}
	\item Intro : positioning between optimization and statistics. Start with stats. Convince the reader that this is an essential problem, with asymptotic rate. but finite is elusive. Inspired by recent work (lepriol2021 , bach2013) we want to use optimization to give a rate. We fail, but we reveal a hole in current analysis.  Finite rate would allow us to quantify the importance of the prior, and also help us understand behavior for small numbers.
	\item Intro : 1 clear question and sharper contributions, with references to sections, as a memo for the reviewer writing his review. "investigate bias-variance..." "we show all these analysis suffer from this problem."
	\item Organization : add optim parallel in insigths section, and then sprinkle references until the optim section.
	\item Optim : find members of the exponential family where the rate of Dragomir et al. holds.
	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
	References: \begin{itemize}
		\item \href{https://www.jstor.org/stable/4616462?seq=1#metadata_info_tab_contents}{Existence of Maximum Likelihood Estimates for Multi-Dimensional Exponential Families},
		\item The singly truncated normal distribution: A non-steep exponential family,
		\item Statistical exponential families: A digest with flash cards,
		\item Information and Exponential Families In Statistical Theory,
		\item Fundamentals of Statistical Exponential Families with Applications in Statistical Decision Theory.
	\end{itemize}
\end{enumerate}

@@

\newpage
\section{Notation convention}
We use
\begin{itemize}
	\item $\theta, \nat$ for natural parameters
	\item $\m$ for mean parameters
	\item $X_1,\ldots,X_n$ for data
	\item $\cX$ for the data space
	\item $T$ for the sufficient statistics
	\item $\nu$ for the base measure
	\item $A, A^*, \logpart, \conj$ for the log-partition and its dual
	\item $\cB_A$ for the Bregman divergence induced by $A$
	\item $n$ for the number of samples
	\item $(\theta^*,m^*)$ for the optimum
	\item $\gamma, \lr$ for the step-size
	\item $\mu,\sigma^2$ for the parameters of a Gaussian
	\item $\hat\theta_n$ for estimates
	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
	\item $(L, \mu)$ for strong-convexity and smoothness
\end{itemize}

\fdk{
	Do we use $\alpha, \beta$ anywhere?
	Might use those for smoothness/strong-convexity rather than $\mu, L$.
}

\fdk{
	The superscript ${}^*$ denotes the convex conjugate when applied on a function
	and the optimum when applied to a parameter.
	How about we use $\theta_n, \theta_*$ for parameter at time $n$ and at the optimum
	and $f^*$ for the convex conjugate?
}

\fdk{
	On $X_i$ for data; I'm more used to seeing $x_i$ for data, which is the realization of the random variable $X_i$.
	We don't talk about random variables and don't need the distinction?
}

\fdk{
	Need to check
}

>>>>

%\textcolor{red}{This comes a bit from nowhere. Maybe phrase it like (its not sexy wirting here, just the idea of the structure) "The measure of closeness between theta and its estimate hat theta is an active area of research. We distinguish two way to do that: 1) distance in the parameter space and 2) distance between their associated distribution. In the first case, the analysis use norms blabla. In the second, we use KL"}
{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality:
how close is it from some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish two ways of doing that.
{\bf 1)} distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$. %Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction} , the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
Hence the fundamental problem:
%\textcolor{red}{The sample complexity of the quality of the estimator is well known when we use norms between parameters, in opposition to the case where we measure the distance between distribution. }
% RLP : that is not so true. the citation tsybakov does gie bound for non-parametric estimation with kernels in L^2 distance.
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We have asymptotic solutions (\S\ref{ssec:asymptote}),
or when $\logpart$ is quadratic -- e.g. $X$ is Gaussian with known variance --
or close to quadratic (see \S\ref{ssec:quadratic}),
but a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions, and give some ideas on how to solve it.
%\textcolor{red}{Maybe say that "in this paper, we show that this problem is difficult and we give the reason why and some idea on how to solve it"}
%RLP problem : we don't show it is difficult, neither do we explain why it is difficult. That sounds hard to do.

@@

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analysis apply, which frames another open problem for the analysis of SMD.

@@

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance})
	\item By identifying MAP and stochastic mirror descent (SMD), we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}

@@

@@

\paragraph{Optimization}
We can rewrite~\eqref{eq:defMAP} as
\alignn{
\m_n = \m_{n-1}- \lr_n (\m_{n-1} - T_n)
\label{eq:mean_update}
}
where $\lr_n := \inv{n_0 + n}$.
Now let $f_X(\nat) = -\log p(X \cond \nat)$ such that $\E[f_X] = f$.
If we further introduce $g_n(\nat) := \nabla\logpart(\nat) - T_n = \nabla f_{X_n}(\nat)$, then~\eqref{eq:mean_update} becomes
\alignn{
	\nabla\conj(\hat \nat_{n})
	= \nabla\conj(\hat \nat_{n-1}) - \lr_n g_n(\hat \nat_{n-1})
}
which is the update formula for stochastic mirror descent (SMD) on $f$ with mirror map $\nabla\logpart$
and step-size schedule $\lr_n$, initialized at $\nat_0$.
\emph{MAP is an instantiation of SMD.}
In this view, MLE forgets its (arbitrary) initialization after the first step with step-size 1.

@@

In theorems~\ref{thm:varianceMLE} and~\ref{thm:varianceMAP}, we report upper bounds on the expected value of this divergence for the MLE and the MAP.
Proofs are in \cref{app:gaussian-variance}.
\begin{theorem}[MLE Tight Bound]
\label{thm:varianceMLE}
	The MLE of $\cN(0,\m_*)$ is $\hat \m_n^\text{MLE} = \inv{n} \sum_i X_i^2 $.
	Its expected suboptimality is infinite when $n\leq 2$, and otherwise upper-bounded as
	\begin{align}
		 \expect{\bregmanconj( \m_*; \hat \m_n^\text{MLE}) }
			\leq \inv{2n} +\frac{2}{n(n-2)} \; .
			\label{eq:MLE_rate}
	\end{align}
\end{theorem}
This upper bound is asymptotically tight.
We illustrate its numerical behavior in Figure~\ref{fig:curves}.
We get a similar bound for the multivariate generalization :
the expected value is infinite whenever $n \leq d+1$ where $d$ is the dimension, and is otherwise bounded by $O(\frac{d^2}{n} + \frac{d^3}{n(n-d-1)} )$ (c.f. App.\ref{app:gaussian-variance}).

@@

\begin{theorem}[MAP Bound]
\label{thm:varianceMAP}
For $n\in \naturalnumbers$, the expected suboptimality of the MAP of $\cN(0,\m^*)$ with prior hyper-parameters $(n_0,\m_0)$ is
 \begin{align}
	& \expect{\bregmanconj( \m_*; \hat \m_n^\mathrm{MAP})}
	\leq \begin{cases}
		\inv{2(n_0+1)}  +  b_1 \ \text{if}\ n=1,\\
		\frac{1}{n_0 \frac{\m_0}{\m^*} +n-2} + b_n \ \text{if}\ n\geq 2
	\end{cases}
	\label{eq:MAP_rate}\\
	& \text{where }b_n = \frac{(1 + \inv{n_0} - \frac{\m_0}{\m^*})^2}{2 (\frac{\m_0}{\m^*}+\frac{\max(0,n-2)}{n_0})(1 + \frac{n}{n_0} )} \; . \nonumber
\end{align}
\end{theorem}
\rlp{upper bound and compare the symmetrized bregman instead?}
This inequality highlights a clear variance-bias decomposition.
In particular, there is no bias term when $\frac{\m_0}{\m^*} =1 + \inv{n_0} $, which happens when the prior is slightly larger than the ground truth.  For instance, when $n_0=1$, it encourages us to set $\m_0 = 2 \m^*$. This correlates well with numerical observations (cf App.\ref{app:gaussian-variance}).
Remark that in using~\eqref{eq:log_bound} we lost a factor $2$ and got a rate$\inv{n}$ instead of $\inv{2n}$.

@@

\subsection{Asymptote}
\label{ssec:asymptote}
As a reference point for any finite convergence rate, it is interesting to know the asymptotic behavior of these quantities as $n \rightarrow +\infty$.
We now review classical asymptotic results.
We point the reader towards \citet[\S1.1]{ostrovskii2021finite} for a more comprehensive overview.
Asymptotic results typically show that the MLE expressed with natural parameters $\nat$ is asymptotically normal: it reaches the Cramér-Rao lower bound \citep[for instance Ch4.2]{vdv1998asymptotic}.
With exponential families, the MAP is more simply expressed with mean parameters $\meanp$.
We get the asymptote by approximating the Bregman divergence using a second order Taylor expansion
\alignn{
    \bregmanconj(\m^* ; \m)
    &= \frac{\norm{\m^* - \m}^2_{\mF}}{2}
    + O(\norm{\m - \m^*}^3),
    \label{eq:bregmanTaylor}
}
where the Mahalanobis norm  $\| x \|_{\mF}^2 = x^\top \mF x$  is induced by $\mF  := \nabla^2\conj(\m^*)$, the hessian of the entropy at the optimum. It happens that  $\mF$ is also the inverse \textit{Fisher information matrix} at $\nat^*$, since
\aligns{
    \mF
    :=\nabla^2\conj(\m^*)
    = \nabla^2\logpart(\nat^*)^{-1}
    = \Cov_{\nat^*}[T(X)]^{-1}  \; .
}
Plugging~\eqref{eq:defMLE} or~\eqref{eq:defMAP} into~\eqref{eq:bregmanTaylor}, one can show that MLE and MAP verify
\begin{align}
	\label{eq:MAP_asymptote}
	\E \bregmanconj \left (\E [T(X)] ; \hat \meanp_n^\text{MLE/MAP} \right )
	= \frac{d}{2n} + O(n^{- \frac{3}{2}}) \; .
\end{align}
Detailed derivation is in~\cref{app:asymptote}
Both MLE and MAP have the same asymptote, as the contribution of the prior $n_0 \meanp_0$ gets negligible compared to the data for large $n$.
This asymptote is also independent of the optimum $\meanp^*$.

@@

\subsection{Locally Quadratic Case}
\label{ssec:local-quadratic}
FromTaylor expansion~~\eqref{eq:bregmanTaylor},
we know that all Bregman divergences are locally quadratic.
Under some assumptions, such as self-concordance of $\conj$\footnote{
In 1d, $f$ is self-concordant iff $\forall x, \abs{f'''(x)} \leq 2 \abs{f''(x)}^{\frac{3}{2}}$.
} \citep[Ch.4.1]{nesterov2003introductory}, we can quantify when this quadratic behavior kicks in. Proofs for this subsection are in \cref{app:self-concordant}.
\begin{proposition}
Let $\conj:\cM\rightarrow \real$ be a self-concordant convex function, $\m, \m^* \in\cM$ and $\mF = \nabla \conj(\m^*)$. Then
\aligns{
	\norm{\m^*-\m}_{\mF} < 0.21
	\implies
	\bregmanconj(\m^*,\m) \leq \norm{\m^*-\m}_{\mF}^2
}
\end{proposition}
\footnote{$0.21$ is a value of $x$ such that $x^2 \geq -\frac{x}{1-x} - \log(1 - \frac{x}{1-x})$}.
To gain insights into how many samples are needed, we can estimate when $\expecti{\norm{\m^*-\m}_{\mF}} < 0.21 $.
For the MLE, $\expecti{\norm{\m^*-\hat \m_n}_{\mF}} = \frac{d}{n}$, so a sufficient condition is $n \geq 5 d$.
For MAP, using~\eqref{eq:MAP_quadratic}, we get the sufficient condition $n\geq 5d + \sqrt{5}\norm{\m^* -  \m_0} - n_0$.
This means that on average, we need much more samples than the dimension to reach the quadratic regime.

@@

\newpage
\bibliographystyle{apalike}
\bibliography{references}

\clearpage
\appendix
\onecolumn

\section{PROOFS FOR GAUSSIAN VARIANCE}
\label{app:gaussian-variance}

\section{ASYMPTOTIC DERIVATION}
\label{app:asymptote}

\section{SELF-CONCORDANCE}
\label{app:self-concordant}

\section{TODO}
\begin{enumerate}
	\item new plots : add 2d trajectories to paper, with background level set, and new legend : primal/dual expectation, and clear explanation !
	\item Incorporate related work by Frederik.
	\item Technical Background : be precise and add reference for the steep exponential family of Legendre type which guarantees that our sets are open, and have a solution in the middle.
	References: \begin{itemize}
		\item \href{https://www.jstor.org/stable/4616462?seq=1#metadata_info_tab_contents}{Existence of Maximum Likelihood Estimates for Multi-Dimensional Exponential Families},
		\item The singly truncated normal distribution: A non-steep exponential family,
		\item Statistical exponential families: A digest with flash cards,
		\item Information and Exponential Families In Statistical Theory,
		\item Fundamentals of Statistical Exponential Families with Applications in Statistical Decision Theory.
	\end{itemize}
\end{enumerate}

@@

\newpage
\section{Notation convention}
We use
\begin{itemize}
	\item $\theta, \nat$ for natural parameters
	\item $\m$ for mean parameters
	\item $X_1,\ldots,X_n$ for data
	\item $\cX$ for the data space
	\item $T$ for the sufficient statistics
	\item $\nu$ for the base measure
	\item $A, A^*, \logpart, \conj$ for the log-partition and its dual
	\item $\cB_A$ for the Bregman divergence induced by $A$
	\item $n$ for the number of samples
	\item $(\theta^*,m^*)$ for the optimum
	\item $\gamma, \lr$ for the step-size
	\item $\mu,\sigma^2$ for the parameters of a Gaussian
	\item $\hat\theta_n$ for estimates
	\item $\tilde\theta_n$ for $\expect{\hat\theta_n}$, which is biased
	\item $(\stgcvx , \smooth)$ for strong-convexity and smoothness
\end{itemize}
